---
title: "p8105_hw3_yg2964"
output: github_document
date: "2024-10-16"
---
```{r}
library(tidyverse)
library(ggridges)
library(patchwork)


library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1
### Section 1: Load the dataset

```{r}
data("ny_noaa")
```

### Section 2: short description of the dataset
The NY NOAA dataset contains `r format(nrow(ny_noaa), big.mark=",")` observations and `r ncol(ny_noaa)` variables. Key variables include weather station ID (id), date of observation, precipitation (prcp, in tenths of mm), snowfall (mm), snow depth (mm), maximum temperature (tmax, in tenths of degrees Celsius), and minimum temperature (tmin, in tenths of degrees Celsius).

### Section 3: Below we clean the data, creating separate variables for year, month, and day and converting tmax and tmin to numeric. We also examine the most commonly observed snowfall values:
```{r}
ny_noaa_clean = 
  ny_noaa |> 
  separate(date, into = c("year", "month", "day"), convert = TRUE) |> 
  mutate(
    tmax = as.numeric(tmax) / 10,  # Convert to Celsius
    tmin = as.numeric(tmin) / 10   # Convert to Celsius
  )

ny_noaa_clean |> 
  count(snow) |> 
  arrange(desc(n))
```

We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is NA, indicating missingness. Other common values are 25, 13, and 51 mm, suggesting that snowfall might have been originally recorded in fractions of an inch (1 inch = 25.4 mm) and converted to mm.

### Section 4: Below is a two-panel plot showing the average max temperature in January and in July in each station across years:
```{r}
ny_noaa_clean |> 
  group_by(id, year, month) |> 
  filter(month %in% c(1, 7)) |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |> 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + 
  geom_point() + 
  geom_path(alpha = 0.3) +
  facet_grid(~month, labeller = labeller(month = c("1" = "January", "7" = "July"))) +
  labs(title = "Mean monthly max temperature for each station across years",
       x = "Year", y = "Mean Max Temperature (°C)")
```
As expected, the mean temperature in January is much lower than in July for all stations across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years. We can observe one uncharacteristically cold station in July of 1988, as well as a few other less drastic outliers. There's also a slight upward trend in temperatures over the years, which could be indicative of climate change.

### Section 5: Below we show a two-panel plot including (i) a hex plot of tmax vs tmin for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values greater than 0 and less than 100 mm separately by year.

```{r}
hex = 
  ny_noaa_clean |>  
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()
  


ridge = 
  ny_noaa_clean |> 
  filter(snow < 100, snow > 0) |> 
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()
hex + ridge
```

From the hex plot, we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution, showing a strong positive correlation between maximum and minimum temperatures. 
From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and 35 mm of snow in a year. Then there is another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm. This multimodality likely stems from the conversion of measurements from inches to millimeters, which was also noted in the table of common values.

## Problem 2
### Section 1:  Data Preparation

```{r}
# Load necessary packages
library(dplyr)
library(janitor)
library(readr)

# Load and clean demographic data
demographics_data <- read_csv("nhanes_covar.csv", na = c("NA", "", ".", " "), skip = 4) |> 
  janitor::clean_names() |> 
  drop_na() |> 
  mutate(education = as.character(education), sex = as.character(sex)) |> 
  mutate(
    education = case_match(
      education,
      "1" ~ "Less than high school",
      "2" ~ "High school equivalent",
      "3" ~ "More than high school"),
    sex = case_match(
      sex,
      "1" ~ "Male",
      "2" ~ "Female"))

# Load and clean accelerometer data
accelerometer_data <- 
  read.csv("nhanes_accel.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names()

# Merge the datasets based on participant ID (corrected)
accel_demo <- inner_join(accelerometer_data, demographics_data, by = "seqn")

# Filter participants by age and encode factors for sex and education
accel_demo <- accel_demo %>%
  filter(age > 21) %>%
  mutate(
    sex = factor(sex, levels = c("Male", "Female")),
    education = factor(education, levels = c(
      "Less than high school", 
      "High school equivalent", 
      "More than high school")
    )
  )
```


### Section 2: 
#### 2.1 Create a table for the number of men and women in each education category
```{r}
education_sex_table <- accel_demo |> 
  group_by(education, sex) |> 
  summarise(count = n()) |> 
  pivot_wider(names_from = sex, values_from = count)
  knitr::kable(education_sex_table)
```

Interpretation of the Table:
From the table, we can observe the following distribution of men and women across education levels:
Less than high school: There are 27 males and 28 females, making the distribution quite balanced in this category.
High school equivalent: There are 34 males and 23 females, indicating a higher proportion of males in this category.
More than high school: There are 54 males and 59 females, making this the largest group overall, with a slightly higher proportion of females.

#### 2.2 Visualization of Age Distributions by Sex and Education:

```{r}
age_sex_plot <- 
  accel_demo |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.6) + 
  labs(
    title = "Age Distributions for Men and Women by Education Level",
    x = "Age",
    y = "Density",
    fill = "Sex"
  ) +
  facet_wrap(~ education) +  
  theme_minimal() +
  theme(legend.position = "bottom") 

age_sex_plot
```
Interpretation:
- Less than High School: Men are concentrated around ages 60-70, while women have a more even spread with peaks around ages 50 and 70.
- High School Equivalent: Men have two peaks, around ages 40-50 and 70. Women are concentrated around ages 50-60.
- More than High School: Men peak around ages 40-50, while women have a younger distribution, peaking around 30-40 and 50-60.
In summary, women tend to have a broader age distribution across education levels, while men are more concentrated in specific age ranges, especially in higher education categories

#### Section 3
```{r}
# Calculate total daily activity for each participant
accel_demo_total <- accel_demo |> 
  rowwise() |> 
  mutate(total_activity = sum(c_across(starts_with("min")), na.rm = TRUE)) |> 
  ungroup()

# Create the plot
total_activity_plot <- ggplot(accel_demo_total, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ education) +
  labs(title = "Total Daily Activity vs Age by Sex and Education Level",
       x = "Age",
       y = "Total Daily Activity (MIMS)",
       color = "Sex") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(total_activity_plot)
```
Interpretation: 
Less than High School:
Total activity decreases with age for both sexes.
Women generally have higher activity levels than men, especially between ages 30-60.

High School Equivalent:
Women show higher total activity than men between ages 30-50.
Both sexes experience a drop in activity after age 60.

More than High School:
Men and women have similar activity levels until age 50, after which women's activity tends to decline more slowly.

#### Section 4:  Create the three-panel plot showing the 24-hour activity time courses for each education level
```{r}
long_activity_data <- accel_demo %>%
  pivot_longer(cols = starts_with("min"), 
               names_to = "minute", 
               values_to = "activity_level") %>%
  mutate(minute = as.numeric(gsub("min", "", minute)))  # Extract minute from variable names

ggplot(long_activity_data, aes(x = minute, y = activity_level, color = sex)) +
  geom_smooth(se = FALSE) +  # Smooth trend line
  facet_wrap(~ education) +  # Separate panels for each education level
  theme_minimal() +
  labs(title = "24-Hour Activity Time Course by Education Level and Sex",
       x = "Minute of Day (0-1440)",
       y = "Activity Level (MIMS)",
       color = "Sex")
```

