---
title: "p8105_hw3_yg2964"
output: github_document
date: "2024-10-16"
---
```{r}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1
## Load the dataset

```{r}
data("ny_noaa")
```

## short description of the dataset
The NY NOAA dataset contains `r format(nrow(ny_noaa), big.mark=",")` observations and `r ncol(ny_noaa)` variables. Key variables include weather station ID (id), date of observation, precipitation (prcp, in tenths of mm), snowfall (mm), snow depth (mm), maximum temperature (tmax, in tenths of degrees Celsius), and minimum temperature (tmin, in tenths of degrees Celsius).











